Project Report\\\textsc{DEVS Ex Machina}

### 2015-05-23 manual/projectrapport.tex
* updated some documentation for the tracing module
Signed-off-by: Lttlemoi <lttlemoi+git@gmail.com>
projectreport + how to build changes

\section{Messaging Size}
A message is constructed once, and after that never copied. Messages are sent and received by reference to smart pointer, incurring a size cost of (on 64-bit platforms) 128bit. The payload (and its size) is irrelevant to the simulation up to the point where the message is actually received. A default implementation allows the payload stored as a std::string, but a specialized message class can be used for sending objects of a different type. When constructing a message by using the functionality provided by the Port class, this happens automatically. When receiving messages, the free function template is provided to extract the object from the message.\\
For example:
\begin{lstlisting}[language=C++,basicstyle=\small\color{black}]
//initialization of necessary objects is left out.
double value;
n_model::t_portptr port;

@@

//Assuming that no Z function was applied to the message.
assert(value == data);
\end{lstlisting}

@@

\section{Translating Dynamic Typing to Static Typing}
To represent objects that can be at runtime any type chosen by the user, we had 2 options : void* which is highly fragile in use, and impossible to serialize, or std::string(), which can double as an "anytype" by storing a serialized object inside the string, should this be required. \\This leaves the same degree of freedom dynamic typing grants to the user in PyPDEVS, with the added benefit of some checking provided by the compiler. Consider for example a user who stores an object serialized into the payload string of a message object, the compiler can in advance detect if the serializing object is at least serializable. \\ Misuse would either surface during compilation, or at worst could be checked with casting or exceptions. Misuse of void* could result in anything from undefined behaviour to stack corruption.

@@

\section{Scheduler}
A Scheduler was implemented based on the prototyping code. As required, as list based scheduler was implemented. \\
The interface from PyPDevs' scheduler class was reused, but internally the scheduler class in our project differs on a few point. \\
\subsection{Composition}
All schedulers are based on a heap, which order items to be scheduled. We never actually schedule models, only entries (A struct containing name and time to schedule), allowing for far faster operation. This is complemented by a hashmap, so that entries can have O(1) lookup and erasure (required for the use case of antimessaging, among others). The operation massReschedule() was left out of the interface, since it can be easily composed using erase() and push\textunderscore back().
\\
\subsection{Implementation}
The actual heap used is either based on std::list as required, or one of Boost's heaps. The user is provided with a convenience factory function that creates Scheduler instances with varying heap backends. In general, the Fibonacci heap is one of the fastest, with skew heap sometimes coming out ahead. The list based heap is always a factor 2 slower, though again this depends on the usage scenario. \\Important to note is that only node based heaps can be used efficiently (without compromising the performance of erase() and contains()). We store on insertion an iterator with the key of the pushed element in the hashmap. This also works as a safety check, preventing double insertion (depending ofcourse on the hash-specialization semantics). This freedom of implementation is visible in the tracers where duplicate elements are allowed if they have a different address, contrasted with the model scheduler which disallows a model to be scheduled twice.\\
Finally, the std::hash$<>$ specialization used by the message scheduler allows O(1) annihilation of a message by its antimessage (since the hash is identical).
\subsection{Differences}
We intended to use the scheduler class not only for scheduling models, but also to store pending incoming messages, and tracing messages. To allow this, the Scheduler class only requires from a scheduled item that it has operator$<$ and std::hash$<$T$>$ defined.

@@

\subsubsection{Teamwork}
We met at least once every week (barring the exam period), with each meeting having a formal report. Whenever required, teammembers used video conferencing to solve critical bugs or discuss design/coding issues.
\subsubsection{Code review}
Although there was no real formal code review done up to now, we've had several debug sessions reviewing each others code. This means the most critical used sections of the code are allready reasonably well reviewed, albeit not formally.
\subsubsection{Portability}
We develop on Unix, Windows and MacOSX.

In Windows we chose the Cygwin platform instead of MinGW to compile our project on. This because we make use of the C++11 std::thread objects; MinGW doesn't support POSIX threads. When building the project with GTest on Windows you need to use the -std=gnu++11 flag instead of the standard -std=c++11 flag, because we need the GNU extensions to correctly link everything.

@@

libstdc++'s std::string implementation appears to be still Copy On Write (despite 21.4.1 p6 from the C++ standard) \cite{cpp}. This is triggers an alert by threadsanitizer for a possible race whenever a string was supposedly copied but internally still stored as a reference to the same object. Explicitly invoking the string constructor with the char[] of the original string, proved to be a safe workaround here. Note that this problem is a known issue \cite{cow}, which is solved in GCC 5.1.\\

@@

	\item No logging code is executed whatsoever if the logger is disabled
	\item log messages contain filename and line number of the log command

@@

@@

\section{Logger}
We have also conducted a small benchmark to compare the raw power of our logger and the proposed g3log.
We let each logger print several hundred thousand messages to a file. These are the performance statistics:
\subsubsection{g3log}
\lstinputlisting[]{g3log.perf}
\subsubsection{Our own logger}
\lstinputlisting[]{own.perf}
\subsubsection{Conclusion}
While the difference is considerable, we should keep in mind that g3log generates timestamps for each message. It also contains several features that our logger doesn't, such as extensive crash safety. Moreover, g3log allows the user to dynamically change which log levels are filtered, while this can only be decided at compile time for our logger. This can be clearly seen in the amount of branches.

@@

>>>>

\section{Messaging Size}
A message is constructed once, and after that never copied. Messages are sent and received by reference to smart pointer, incurring a size cost of (on 64-bit platforms) 128bit. The payload (and its size) is irrelevant to the simulation up to the point where the message is actually received. A default implementation allows the payload stored as a std::string, but a specialized message class can be used for sending objects of a different type. When constructing a message by using the functionality provided by the Port class, this happens automatically. When receiving messages, a free function template is provided to extract the object from the message.\\
For example:
\begin{lstlisting}[language=C++,basicstyle=\small\color{black}]
//initialization of necessary objects is left out.
double value;
n_model::t_portptr port;

@@

//Assuming that no Z function was applied to the message.
assert(value == data);
\end{lstlisting}
\paragraph{Note} While it is technically possible, we strongly discourage the usage of the following data types in these messages:
\begin{itemize}
	\item raw pointers to heap-allocated object.\\
			Messages can be destroyed during a time warp prior to being delivered to a model. This will result in a memory leak if there are no other references to these objects.
	\item pointers or references to local objects.\\
		When these objects go out of scope, they are destructed, rendering the pointer or reference invalid.
\end{itemize}

@@

\section{Translating Dynamic Typing to Static Typing}
Certain classes of the simulation, such as the messages between models and the model states, must be able to contain objects of any type. However, C++ is a statically typed language. To overcome this problem, we came up with the following three solutions:
\begin{itemize}
	\item Keep a \texttt{void*} that points to the actual data. This approach is very fragile and impossible to serialize. Misuse of \texttt{void*} could result in anything from undefined behaviour to stack corruption.
	\item Serialize everything to \texttt{std::string}. Serialization is not a trivial operation. Continuously serializing and deserializing can be a very large performance hit. Moreover, it would require each and every type to be serializable. On the flip side, the compiler can help by detecting whether or not the object can be serialized or not.
	\item Make use of inheritance and create a specialized subclass for each and every datatype. In the case of the messages, a class template can be used to facilitate this process. In the case of the model states, we decided that it is best that the user provides the subclass.\\
	Since we use pointers for messages and states, a pointer cast is all that is needed to access the members of the specialized class. For extra safety, we suggest to use \texttt{std::dynamic\_pointer\_cast}.
\end{itemize}
We chose the third option because it both offers speed and safety.

@@

\section{Scheduler}
A Scheduler was implemented based on the prototyping code. As required, as list based scheduler was implemented. \\
The interface from PyPDevs' scheduler class was reused, but internally the scheduler class in our project differs on a few points. \\
\subsection{Composition}
All schedulers are based on a heap, which order items to be scheduled. We never actually schedule models, only entries (A struct containing name and time to schedule), allowing for far faster operation. This is complemented by a hashmap, so that entries can have O(1) lookup and erasure (required for the use case of antimessaging, among others). The operation massReschedule() was left out of the interface, since it can be easily composed using erase() and push\textunderscore back().
\\
\subsection{Implementation}
The actual heap used is either based on std::list as required, or one of Boost's heaps. The user is provided with a convenience factory function that creates Scheduler instances with varying heap backends. In general, the Fibonacci heap is one of the fastest, with skew heap sometimes coming out ahead. The list based heap is always a factor 2 slower, though again this depends on the usage scenario. \\Important to note is that only node based heaps can be used efficiently (without compromising the performance of erase() and contains()). We store on insertion an iterator with the key of the pushed element in the hashmap. This also works as a safety check, preventing double insertion (depending ofcourse on the hash-specialization semantics). This freedom of implementation is visible in the tracers where duplicate elements are allowed if they have a different address, contrasted with the model scheduler which disallows a model to be scheduled twice.\\
Finally, the std::hash$<>$ specialization used by the message scheduler allows O(1) annihilation of a message by its antimessage (since the hash is identical).
\subsection{Differences}
We intended to use the scheduler class not only for scheduling models, but also to store pending incoming messages, and tracing messages. To allow this, the Scheduler class only requires from a scheduled item that it has operator$<$ and std::hash$<$T$>$ defined.

@@

\subsubsection{Teamwork}
We met at least once every week (barring the exam period), with each meeting having a formal report. Whenever required, teammembers used video conferencing to solve critical bugs or discuss design/coding issues.
\subsubsection{Code review}
Although there was no real formal code review done up to now, we've had several debug sessions reviewing each others code. This means the most critical used sections of the code are allready reasonably well reviewed, albeit not formally.
\subsubsection{Portability}
We develop on Unix, Windows and MacOSX. In Windows we chose the Cygwin platform instead of MinGW to compile our project on. This because we make use of the C++11 std::thread objects; MinGW doesn't support POSIX threads.

@@

libstdc++'s std::string implementation appears to be still Copy On Write (despite 21.4.1 p6 from the C++ standard) \cite{cpp}. This triggers an alert by threadsanitizer for a possible race whenever a string was supposedly copied but internally still stored as a reference to the same object. Explicitly invoking the string constructor with the char[] of the original string, proved to be a safe workaround here. Note that this problem is a known issue \cite{cow}, which is solved in GCC 5.1.\\

@@

	\item No logging code is executed whatsoever if the logger is disabled.
	\item log messages contain file name and line number of the log command.

@@

\paragraph{Asynchronous output}
To implement the asynchronous output, we made use of a custom stream buffer. The original design is by Dietmar K{\"u}hl \cite{asynchwrite}. We modified this design in order to make it thread safe.

\section{Compiling multi-platform}
A lot of obstacles were tackled while getting the project to work an Windows. When building with GTest on Windows you need to use the \texttt{-std=gnu++11} flag instead of the standard \texttt{-std=c++11} flag, because we need the GNU extensions to correctly link everything.

All standard c++11 \textit{to string} functionality in Cygwin is not supported \cite{cygwin_tostring} (MinGW32 doesn't support this either). With basic \textit{to string} functionality we mean the functions that are listed here \cite{c++11_tostring} and under the 'See also' section. The problem seems to be a library called Newlib \cite{newlib} that doesn't provide crucial functions.

Mac OSX needs the libc++ library to compile all C++11 features on.

@@

\section{Logger}
We have also conducted a small benchmark to compare the raw power of our logger and the proposed g3log.
We let each logger print several hundred thousand messages to a file. These are the performance statistics:
\subsubsection{g3log}
\lstinputlisting[]{g3log.perf}
\subsubsection{Our own logger}
\lstinputlisting[]{own.perf}
\subsubsection{Conclusion}
While the difference is considerable, we should keep in mind that g3log generates timestamps for each message. It also contains several features that our logger doesn't, such as a more extensive crash safety. Moreover, g3log allows the user to dynamically change which log levels are filtered, while this can only be decided at compile time for our logger. This can be clearly seen in the amount of branches taken by the program.

@@

  \bibitem{asynchwrite} \url{http://stackoverflow.com/a/21127776}

  \bibitem{c++11_tostring} \url{http://en.cppreference.com/w/cpp/string/basic_string/to_string}

  \bibitem{cygwin_tostring} \url{https://cygwin.com/ml/cygwin/2015-01/msg00245.html}

  \bibitem{newlib} \url{https://sourceware.org/newlib/}
